master thesis

### 2018-04-02 Paper - Latex/Paper - Skizze - v4.tex
Thesis - Discussion again

      The Eukaryota synthesis tree has 293,463 leaf nodes. The GloBI database has 5,346,414 interactions 
        (at 29.01.2018). Out of this data we recieved 51,337 free-living species and 47,332 parasite 
        species for the whole tree of life. For the Eukaryota we could determine 25,962 and 34,860 
        species as parasites and free-living. With 2,293,463 leaf nodes we still have about 97.34\% 
        unknown leaf nodes.

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
\chapter{Discussion}
  This chapter deals with the question of how trustworthy our result of the ancestral state 
    reconstruction of the Eukaryota tree is. And further on how well our simulation can simulate the 
    real problem and thus make statements about the predictive power of the applied Sankoff algorithm.

  We have pursued this question in various investigations and yet, of course, further possibilities 
    for improvement remain. Despite these possibilities for improvement, our reconstruction gives a 
    first good assessment of the whole tree.

  In the first part of this work, we were able to provide the proof of concept for the Sankoff 
    algorithm to perform an ancestral state reconstruction of the present Eukaryota tree. In the 
    following, we will first discuss the data situation and the simulation that led to the selection 
    of the Sankoff algorithm, and then discuss how significant the result of this reconstruction is.

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Data situation} \label{sec:discussion - data situation}
    The used Eukaryota synthesis tree from OTL \cite{Hinchliff2015} has 241,974 internal nodes and 
      293,463 leaf nodes. In addition, we could specify 25,962 parasitic and 34,860 free-living 
      species from GloBI \cite{Poelen2014}. \\
    This gives us a high number of missing internal nodes (high multifurcation) and a low number of 
      node states of the leaf nodes in the entire tree.
    We suspect that this lack of information is not equally distributed in both cases. We conclude 
      this from various observations.

    For one thing, there were a few subtrees of higher ranks whose information share was much higher. 
      In the Chordata there is only 66.49\% multifurcation and in the Platyhelminthes only 68.73\% 
      missing state information. \\
    Furthermore, we know that OTL is a synthesis of 819 phylogenetic trees. This suggests that just 
      these parts of the tree are very well resolved and the remainder consists only of the rough 
      taxonomy. For a review of this underlying taxonomy, we have compiled a list of all taxa and 
      plottet them in relation to their distribution (see appendix section \ref{sec:appendix - otl analysis}). 
      There the number of nodes in the tree corresponding to a taxonomic rank is tabulated.

@@

    From GloBI we got very little information about leaf node states compared to their number in the 
      Eukaryota tree. Nevertheless, the prediction of leave-100-out cross-validation was very good. 
      This also leads to the conclusion that the information is very clustered.

    As a last point, we would like to cite the modeling of the missing information. The fact that this 
      is getting better, the lower ordered taxa (smaller subtrees) we are looking at, also suggests a 
      high variability of the data.

    We would like to mention a few other aspects of the GloBI database here. \\
    Of course, such large data sets are flawed. We found some misinformation and were able to report 
      some of these directly to GloBI. \\
    We also found that there is some unused information in GloBI. We therefore found some more source 
      species and target species without OTT identifiers. Since we currently use only OTT 
      identifiers, we could not use this information. At this point there is thus the possibility to 
      use more of the existing data, if one performs a matching with the other identifiers.

@@

    We assume that different parasite types have different transition probabilities. It is therefore 
      difficult to establish a common distribution across the Eukaryota tree. \\
    Based on the estimates of Windsor \cite{Windsor1998}, we have assumed a distribution of 40\% 
      parasites to 60\% free-livings in this work. As a result of the diversity of parasites and the 
      lack of generalizations, we have generally stated that 
      $\mathcal{P}(FL \rightarrow P) = \mathcal{P}(P \rightarrow FL)$. But it is also reasonable to 
      assume that in general $\mathcal{P}(FL \rightarrow P) > \mathcal{P}(P \rightarrow FL)$, because 
      a reverse mutation is usually less likely. However, one would have to determine how much this 
      difference is and thus discuss another parameter.

    In the simulation, we found that this combination has a significant influence on the predictive 
      power of the algorithms. \\
    We have adapted the combination of distributions and threshold to a high probability of achieving 
      the given proportions on initial attempts. However, this means that we prefer high transition 
      probabilities for both transitions and thus the simulated trees have a large number of 
      transitions. It can be assumed that this may not correspond to the real data and thus means a 
      limitation of our simulation. Further we assume that with less number of state changes, the 
      predictive power of both algorithms would be even better, thus this limitation is probably not 
      very restrictive. \\
    At this point it would be possible to test other distributions with equal threshold values. this 
      would be more computationally expensive, but would result in fewer transitions.

@@

    The used \textit{castor} package \cite{Louca2017} offers the possibility to enter different transition 
      probabilities, so with information about these probabilities, you might be able to improve the 
      prediction here.

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Biological view}
    To analyze the results, we have selected some phyla (subtrees) to evaluate our results selectively 
      from the biological point of view: Chordata, Nematoda, Platyhelminthes and Apicomplexa.

    Several factors such as the distribution of existing input data via parasites and free-livings, 
      faulty input data from GloBI and reinforcement of errors by multifurcation play an important 
      role which becomes evident in these examples.

    The accuracy of our results stands and falls with the presence and the correctness of the data of 
      GloBI. Errors of incorrect input data can be amplified by incorrect prediction of unknown 
      species and can be reversed in order to improve the data situation of GloBI. \\
    Since we look at such large trees we can not expect to know all the parasites, so we look at 
      individual positives. They are positive in the sense that the majority have the opposite state.

    In contrast to the other phyla examined, the phylogeny in the Chordata is more pronounced (less 
      multifurcation). This leads to a lower variance of errors, which is reflected in the results. 
      There are 18 parasites as input data and only 8 more are predicted. The Chordata mostly consist 
      of free-living species, so this seems believable. We started with 99.83\%  species and predict 
      99.97\% species as free-living (including already known nodes). \\
    For a detailed view we have created and described a tree with all known and predicted parasites 
      (see appendix \ref{sec:parasites in chordata}).

    The Apicomplexa are a parasitic phylum. We found only one input organism: \textit{Stemonitis fusca} 
      as a free-living species. In GloBI it is listed as being parasitized by Nectria candicans and 
      Nectriopsis sporangiicola\footnote{
        \hyperlink{https://www.globalbioticinteractions.org/?interactionType=parasiteOf&targetTaxon=Stemonitis\%20fusca}
        {https://www.globalbioticinteractions.org/?interactionType=parasiteOf\&targetTaxon=Stemonitis\%20fusca};
        Last checked: 22.03.2018.
      }. The algorithm has not predicted new free-livings.

    Most species of Platyhelminthes (flatworms) are parasites, although there are also free-living, 
      predatory feeding species. These are summarized in the Turbellaria, while the parasites are 
      divided into three other classes \cite{Ax1961}. This also corresponds to our observations. There 
      is one class (Rhabditophora) that contains all but one single exception of free-living species 
      of this phylum, which includes the Turbellaria. \\
    It should be noted, however, that this classification is outdated, as it has been proven that 
      the Turbellaria are not monophyletic. But we will not go into detail here. \\
    For the Platyhelminthes we have more state information for the leaf nodes compared to the other 
      considered subtrees. We start with 0.1\% free-livings and predicted 0.77\% as free-living 
      species.

    With the Nematoda it looks more complicated. Large parts of Nematoda are free-living, but we 
      found only 5.32\% of them. Blaxter and Koutsovoulos estimate the order of 25,000 parasites in 
      the Nematoda and and speak of at 18 independently arosen parasites in Nematoda 
      \cite{Blaxter2015}. \\
    % The problem at this point, however, is: Hallan speaks of the fact that only 23,000 species are 
    %   described by the Nematoda but it is assumed that one million or more species are in this 
    %   phylum.\footnote{J. Hallan, unpublished; 
    %     \hyperlink{https://insects.tamu.edu/research/collection/hallan/}
    %     {https://insects.tamu.edu/research/collection/hallan/}
    %   }. \todo{Link ist nicht erreichbar!} 
    We assume that the parasites have been much more studied and thus we start with only 0.63\% 
      free-living species. Against such a shifted data situation, the algorithm is almost powerless to 
      make correct predictions. And yet the percentage has increased to 5.34\%. \\
    The root node of the Nematoda is predicted as a parasite and so we got 11 losses of parasitism 
      and only 2 origins in this phylum. If we assume that the root node of Nematoda is free-living, 
      then some losses would have to turn around and become Origins. So it could be that we end up 
      with a similar size as Blaxter and Koutsovoulos \cite{Blaxter2015}.

@@

    From the simulation, we can conclude that we predict correctly about 60\% of the nodes in the 
      present data situation. The leave-100-out cross-validation even showed that we predict the 
      omitted nodes to be 98.17\% percent correct. \\
    This allows for the assumption that the data is grouped and not uniformly distributed and thus 
      smaller subtrees are present in which data are to be found in the simulation with smaller 
      multifurcation and smaller value for unknown nodes. This is also confirmed by our statistical 
      and biological analysis of the considered subtrees: Chordata, Nematoda, Platyhelminthes and 
      Apicomplexa. \\
    However, this means that the ancestral states' data in the direction of root node are probably 
      particularly unbelievable. This makes the localization of origins in the direction of the root 
      node difficult. The question remains, how much this affects the estimation of the number of 
      origins. Our comparison with the paper from Weinstein and Kuris \cite{Weinstein2016} an with
      Blaxter and Koutsovoulos \cite{Blaxter2015} showed similar numbers of origins in some subtrees, 
      which leaves us with optimism.

>>>>

      The Eukaryota synthesis tree has 293,463 leaf nodes. The GloBI database has 5,346,414 interactions 
        (at 29.01.2018). Out of this data we received 51,337 free-living species and 47,332 parasite 
        species for the whole tree of life. For the Eukaryota we could determine 25,962 and 34,860 
        species as parasites and free-living. With 2,293,463 leaf nodes we still have about 97.34\% 
        unknown leaf nodes.

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
\chapter{Discussion}
  This chapter is about how trustworthy our result of the ancestral state reconstruction of the 
    Eukaryota tree is and how good our simulation is in order to make statements about the predictive 
    power of the applied Sankoff algorithm.

  We pursued this question in various investigations and yet, of course, further possibilities for 
    improvement remain. Despite these improvement opportunities, our reconstruction gives a first 
    good assessment of the whole tree.

  In the first part of this work, we were able to provide the proof of concept for the Sankoff 
    algorithm to perform an ancestral state reconstruction of the present Eukaryota tree. Hereinafter,

    we will first discuss the data situation and the simulation that led to the selection of the 
    Sankoff algorithm, followed by a discussion about the significance of the result of this 
    reconstruction.

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Data situation}
    The used Eukaryota synthesis tree from OTL \cite{Hinchliff2015} has 241,974 internal nodes and 
      293,463 leaf nodes. In addition, we could specify 25,962 parasitic and 34,860 free-living 
      species from GloBI \cite{Poelen2014}. \\
    This gives us a high number of missing internal nodes (high multifurcation) and a low number of 
      states of the leaf nodes in the entire tree.
    We suspect that this lack of information is not equally distributed in both cases. We conclude 
      this from various observations:

    For one, there were a few subtrees of higher ranks whose information share was much greater. In 
      the Chordata there is only 66.49\% multifurcation and in the Platyhelminthes only 68.73\% 
      missing state information.

    Furthermore, we know that OTL is a synthesis of 819 phylogenetic trees. This suggests that just 
      these parts of the tree are very well resolved and the remainder consists only of the rough 
      taxonomy. For a review of this underlying taxonomy, we have compiled a list of all taxa and 
      plotted them in relation to their distribution (see appendix section \ref{sec:appendix - otl analysis}). 
      In this list, the number of nodes in the tree corresponding to a taxonomic rank is tabulated.

@@

    From GloBI we received very little information about leaf node states compared to their quantity 
      in the Eukaryota tree. Nevertheless, the prediction of the leave-100-out cross-validation was 
      very good. This also leads to the conclusion that the information is very clustered.

    As a last point, we would like to cite the modeling of the missing information. The fact that this 
      gets better, the lower the taxa (smaller subtrees) we are looking at, also indicates a high 
      variability of the data.

    We would like to mention a few other aspects of the GloBI database here. Of course, such large 
      data sets are flawed. We found some misinformation and were able to report some of these 
      directly to GloBI. We also noticed that there is some unused information in GloBI. We found some 
      source species and target species without OTT identifiers. Since we currently use only OTT 
      identifiers, we could not use this information. At this point there is thus the possibility to 
      use more of the existing data, if one performs a matching with the other identifiers.

@@

    We assume that different parasite types have different transition probabilities. It is therefore 
      difficult to establish a common distribution across the Eukaryota tree. Based on the estimates 
      of Windsor \cite{Windsor1998}, we have assumed a distribution of 40\% parasites to 60\% 
      free-livings in this study. As a result of the diversity of parasites and the lack of 
      generalizations, we have stated that 
      $\mathcal{P}(FL \rightarrow P) = \mathcal{P}(P \rightarrow FL)$. But it is also reasonable to 
      assume that in general $\mathcal{P}(FL \rightarrow P) > \mathcal{P}(P \rightarrow FL)$, because 
      a reverse mutation is usually less likely. However, one would have to determine how much this 
      difference is and thus discuss another parameter.

    In the simulation, we found that this combination has a significant influence on the predictive 
      power of the algorithms. We have adapted the combination of distributions and threshold to a 
      high probability of achieving the given proportions on initial attempts. However, this means 
      that we prefer high transition probabilities for both transitions and thus the simulated trees 
      have a large number of transitions. It can be assumed that this may not correspond to the real 
      data and thus means a limitation of our simulation. Furthermore we assume that with a lower 
      number of state changes, the predictive power of both algorithms would be even better, thus this 
      limitation is probably not very restrictive. At this point it would be possible to test other 
      distributions with equal threshold values. This would be more computationally expensive, but 
      would result in fewer transitions.

@@

    The used \textit{castor} package \cite{Louca2017} offers the possibility to enter different 
      transition probabilities, so with information about these probabilities, one might be able to 
      improve the prediction here.

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Biological view}
    We have selected some phyla (subtrees) to evaluate our results selectively 
      from the biological point of view: Chordata, Nematoda, Platyhelminthes and Apicomplexa. Various 
      factors such as the distribution of existing input data about parasites and free-livings, 
      erroneous input data from GloBI, and the amplification of these errors due to multifurcation 
      play an important role, which becomes clear in these examples.

    The accuracy of our results stands and falls with the presence and the correctness of the data of 
      GloBI. Errors that result from incorrect data can be amplified by incorrect prediction of unknown 
      species and can be reversed in order to improve the data situation of GloBI. Since we look at 
      such large trees we can not expect to know all the parasites, so we look at individual positives. 
      They are positive in the sense that the majority have the opposite state.

    In contrast to the other phyla examined, the phylogeny in the Chordata is more pronounced (less 
      multifurcation). This leads to a lower variance of errors, which is reflected in the results. 
      There are 18 parasites as input data and only 8 more are predicted. The Chordata mostly consist 
      of free-living species, so this seems credible. We started with 99.83\%  species and predicted
      99.97\% species as free-living (including already known nodes). For a detailed view we created 
      and described a tree with all known and predicted parasites (see appendix 
      \ref{sec:parasites in chordata}).

    The Apicomplexa are a parasitic phylum. We found only one input organism: \textit{Stemonitis fusca} 
      is a free-living species. It is listed in GloBI as being parasitized by Nectria candicans and 
      Nectriopsis sporangiicola\footnote{
        \hyperlink{https://www.globalbioticinteractions.org/?interactionType=parasiteOf&targetTaxon=Stemonitis\%20fusca}
        {https://www.globalbioticinteractions.org/?interactionType=parasiteOf\&targetTaxon=Stemonitis\%20fusca};
        Last checked: 22.03.2018.
      }. The algorithm has not predicted new free-livings.

    Most species of Platyhelminthes (flatworms) are parasites, although there are also free-living, 
      predatory feeding species. These are summarized in the Turbellaria, while the parasites are 
      divided into three other classes \cite{Ax1961}. This also corresponds to our observations. There 
      is one class (Rhabditophora) that contains all but one single exception of free-living species 
      of this phylum, which includes the Turbellaria. For the Platyhelminthes we have proportionally 
      more state information for the leaf nodes compared to the other considered subtrees. We start 
      with 0.1\% free-livings and predicted 0.77\% as free-living species.

    The Nematoda are more complicated. Large parts of Nematoda are free-living, but we found only 
      5.32\% of them. Blaxter and Koutsovoulos estimate the order of 25,000 parasites in the Nematoda 
      and speak of 18 independently arosen parasites in Nematoda \cite{Blaxter2015}. We assume that 
      the parasites have been much more studied and thus we start with only 0.63\% free-living species. 
      Against such a shifted data situation, the algorithm is almost powerless to make correct 
      predictions. And yet the percentage has increased to 5.34\%.

    The root node of the Nematoda is predicted as a parasite and so we received 11 losses of parasitism 
      and only 2 origins in this phylum. If we assume that the root node of Nematoda is free-living, 
      then some losses would have to turn around and become Origins. So it could be that we end up 
      with a similar size as Blaxter and Koutsovoulos \cite{Blaxter2015}.

@@

    From the simulation, we can conclude that we predict about 60\% of the nodes in the present data 
      situation correctly. The leave-100-out cross-validation even showed that we predict the omitted 
      nodes to be 98.17\% percent correct.

    This allows for the assumption that the data is grouped and not uniformly distributed and thus 
      smaller subtrees are present in which data are to be found in the simulation with smaller 
      multifurcation and smaller value for unknown nodes. This is also confirmed by our statistical 
      and biological analysis of the considered subtrees: Chordata, Nematoda, Platyhelminthes and 
      Apicomplexa.

    However, it follows that the ancestral states' data in the direction of root node are probably 
      particularly unreliable. This makes the localization of origins in direction of the root node 
      difficult. The question remains, how much this affects the estimation of the number of origins. 
      Our comparison with the paper from Weinstein and Kuris \cite{Weinstein2016} an with Blaxter and 
      Koutsovoulos \cite{Blaxter2015} showed similar numbers of origins in some subtrees, which 
      leaves us with optimism.
